\Large{\textbf{A biophysical minimal model to investigate age-related changes in CA1 pyramidal cell electrical activity}}

### 2024-07-23 paper/McKiernan_CA1paper_clean.tex
Update manuscript files

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Aging is a physiological process that is still poorly understood, especially with respect to effects on the brain. There are open questions about aging that are difficult to answer with an experimental approach. Underlying challenges include  the difficulty of recording \textit{in vivo} single cell and network activity simultaneously with submillisecond  resolution, and brain compensatory mechanisms triggered by genetic, pharmacologic, or behavioral manipulations. Mathematical modeling can help address some of these questions by allowing us to fix parameters that cannot be controlled experimentally and investigate neural activity under different conditions. We present a biophysical minimal model of CA1 pyramidal cells (PCs) based on general expressions for transmembrane ion transport derived from thermodynamical principles. The model allows directly varying the contribution of ion channels by changing their number. By analyzing the dynamics of the model, we find parameter ranges that reproduce the variability in electrical activity seen in PCs. In addition, increasing the L-type {\Ca} channel expression in the model reproduces age-related changes in \edit{electrical activity} that are qualitatively and quantitatively similar to those observed in PCs from aged animals. \edit{We also make predictions about age-related changes in PC bursting activity that, to our knowledge, have not been reported previously. We conclude that the model's biophysical nature, flexibility, and computational simplicity make it a potentially powerful complement to experimental studies of aging.}

@@

It is not well understood how changes in ion channel gene expression and hippocampal PC excitability may affect neuron responsiveness and microcircuit output. In part, this is due
to challenges inherent in performing the needed experiments. Single PCs are difficult
to access in intact animals where hippocampal microcircuit function is
preserved. It is also difficult to tease apart the influence of the many different neurophysiological factors that change during aging. Mathematical modeling provides \edit{a} means to understand more about the effects of aging on hippocampal cellular excitability by controlling factors we cannot control experimentally.

@@

\subsection*{Model}
To simulate the electrical activity of CA1 PCs, we used an extended version of a two-dimensional, biophysical model previously developed and characterized by two of the present authors  \cite{herrera2012membranes,herrera2013relating,herrera2018thermodynamic}. \edit{The present model differs from our previous formulations in that it includes \Ca dynamics, i.e. it is three-dimensional, allowing for additional behaviors like bursting \cite{avron1993basic}. In addition, it is specially tuned by incorporating specific channel variants and corresponding experimental data from hippocampal PCs, as described below.} The equations for the ionic currents are derived from first principles of thermodynamics. Previous modeling studies have shown that to reproduce firing behaviors such as spike frequency adaptation and bursting, the minimum number of variables is three \cite{hindmarsh1984model,avron1993basic}. 
In particular, {\Ca} dynamics are important for producing adaptation and burst firing in CA1 PCs (for review see \cite{mckiernan2017ca1}). 
The model dynamics are therefore described by three ordinary differential equations for the time-dependent changes in the transmembrane potential ($v$, in mV), the proportion of open {\K} channels ($w$ in [0,1]), and the intracellular {\Ca} concentration ($c$, in $\mu$M), respectively   \cite{herrera2018thermodynamic}. Based on a well-known relationship between voltage-dependent activation of delayed rectifier {\K} channels and inactivation of {\Na} channels, $w$ also represents the proportion of inactivated {\Na} channels \cite{rinzel1985excitation,avron1991minimal}.

It is assumed that the membrane potential changes due to currents produced by ions transported across the membrane. We take into account currents mediated by a voltage-gated inactivating {\Na} channel ($I_{NaT}$), \edit{voltage-gated L-type {\Ca} channel ($I_{CaL}$),} voltage- and {\Ca}-gated {\K} channels ($I_{DK}$ and $I_{SK}$, respectively), and a {\Na}/{\K}-ATPase ($I_{NaK}$). \edit{We also incorporate a forcing term ($I_F$), which can be used to stimulate the model PC and is explained in more detail below.} The time-dependent change in membrane potential can be written as
\begin{eqnarray}
\label{eq:dvdt}
C_m \partial_t v= I_{F} - I_{NaT}(v,w) - I_{CaL}(v,c) - I_{DK}(v,w) - I_{SK}(v,c) - I_{NaK}(v). 
\end{eqnarray}
with $\partial_t$ representing the instantaneous change with respect to time. $C_m$ (pF) is a constant representing the change in the density of charge around the membrane with respect to voltage, typically referred to as membrane capacitance in models based on electrical circuits  \cite{herrera2020nonequivalent}. Based on recordings from rat CA1 PCs, it is  assumed that $C_m=25$  pF \cite{groc2002vivo}.

@@

For the dynamics of intracellular {\Ca}, we assume recovery toward a steady state $c_{\infty}$ at a rate $r_c$, with increments caused by the {\Ca} current $I_{Ca}$ \cite{avron1991minimal},
\begin{eqnarray}
\partial_t{c} = r_{c}({c_{\infty} - c}) - \tilde{k}_{c} I_{CaL}(v,c).
\label{eq:dcdt}
\end{eqnarray}
The term $\tilde{k}_c$ in equation~\eqref{eq:dcdt} is a conversion factor ($\mu$M/pCoul) that accounts for the effect of {\Ca} flux across the membrane on the intracellular {\Ca} concentration.

The term $I_{F}$ represents a stimulus \textit{forcing} the membrane; that is, current from an electrode \edit{($I_{Stim}$)}, or time-dependent fluctuations from the local field potential \edit{(LFP)}. LFP activity is simulated by replacing the term $I_{F}$ with a  time-dependent, Ornstein-Uhlenbeck (OU) process with amplitude $a_{F}(t)$ (pA). The mean is represented by  $\mu_{F}$ (pA) (drift term)
\cite{rudolph2003characterization} given by \cite{gillespie1996exact} 
\begin{eqnarray}
\label{eq:OU}
{a}_{F}(t+\delta) &=& a_{F}(t) \lrRound{1 -  \frac{\delta}{\tau_{F}}}  +
\lrSquare{ \mu_{F} \delta + \eta(t) ~\sqrt{d_{Stim} \delta}},
\end{eqnarray}
where $\delta$ is a small time step, $\tau_{F}$ is a relaxation time, and $\eta(t)$ is an independent
white noise process with zero-mean and unit standard deviation. \edit{In our simulations,} the mean is set close to the rheobase for the model PCs ($\sim$50 pA). The process has a variance
$\sigma_{F}^2 = d_{F} \delta /2$ (pA), which means $d_{F}$ can be
approximated if an estimation of the variance of the current $a_{F}$ is
available \cite{rudolph2004method,destexhe2004novel}.

@@

The term $J_{F}$ (1/ms) is the input current $I_F$ (pA) divided by $v_T C_m$. After the change in variables and the normalization of the current amplitudes,  equation~\eqref{eq:dcdt} changes to 
\begin{eqnarray}
\label{eq:dcdt2}
\partial_t{c} = r_{c}({c_{\infty} - c}) - k_{c} A_{CaL}n_{\infty}(u
) \varphi_{Ca}(u,c),
\end{eqnarray}
where $k_c = \tilde{k}_c v_T C_m$.

@@

By exploring the model through parameter variations, we were able to find parameter sets that produced different firing patterns, such as adaptive firing, conditional bursting, and spontaneous bursting. \edit{The rationale behind finding parameters for qualitatively different firing patterns that emerge in a three-dimensional model similar to this one was established by Av-Ron and colleagues \cite{avron1993basic}. For example, they show that one way to observe transitions between adaptive firing and bursting is to vary the ratio of DK to \Ca-dependent \K channels. Therefore, in our study, we began by setting the model parameters within the base physiological range given by experimental recordings in PCs, and then took into consideration the different ratios of select parameters and their effects. Once we defined these different restricted parameter ranges, we then tuned semi-manually, i.e. using Python for loops to quickly run through a series of parameter values using knowledge of how increases or decreases in specific parameters should change the firing of cells (e.g., a larger SK current leads to greater inhibition of firing, lower firing frequencies, etc.).} The \edit{specific PC} firing patterns are described in more detail in the Results section, but the respective parameter sets are included in Table \ref{tab:regimes} for ease of comparison.

@@

\subsection*{Simulations}
All code was written in Python 3.7.4 and run on MacBook Pro laptops with 2.9 GHz Intel Core i5 processors. Simulations were performed using functions from the Python library NumPy \cite{numpyHarris2020}. Figures were produced with the Python library Matplotlib \cite{hunter2007matplotlib}. OU processes were simulated using the pyprocess \edit{package} \cite{mondaca200PyProcess}.

\subsection*{Resource availability}
Resources generated by this study (code, figures, and manuscript files) are available via GitHub (\href{https://github.com/emckiernan/agingCA1}{github.com/emckiernan/agingCA1}) and archived via Zenodo (\href{https://doi.org/10.5281/zenodo.6788229}{doi.org/10.5281/zenodo.6788229}). To facilitate reuse, resources are shared under open licenses (see the license file in our GitHub repository). To promote reproducibility, Python code is embedded in a Jupyter notebook \cite{kluyver2016jupyter} that explains the code, how to use it, and how to generate \edit{all} figures, \edit{including those in the Supporting Information}.

\subsection*{Study design}
While aged cells display a number of biophysical changes, we focused on their {\Ca} channel expression. Aged CA1 PCs show an increase in the number of functional transmembrane L-type {\Ca} channels  \cite{herman1998up,thibault1996increase,veng2002regionally}. In particular, CA1 PCs from aged rats have increased expression of Ca$_v$1.2 at the plasma membrane \cite{nunez2014surface}. With these results in mind, we decided to simulate one aspect of aging by changing the number of Ca$_v$1.2 channels in our model membrane. We asked the question, is a change in Ca$_v$1.2 expression sufficient to reproduce the various changes in excitability, such as increased spike frequency adaptation, observed experimentally in aged CA1 PCs? In addition, CA1 PCs are known to burst \cite{mckiernan2017ca1}, but we are not aware of studies comparing \edit{their} bursting patterns in young versus aged animals. Therefore, we used our model to also investigate the effects of altered Ca$_v$1.2 channel expression on bursting activity. In all the following simulations, once the models have been tuned to a specific firing pattern, young and aged model PCs (hereafter referred to as yPC and aPC, respectively) are identical with respect to every parameter except the maximum amplitude of their L-type {\Ca} current, which is set to produce currents of $\sim$2-3 nA or $\sim$5-6 nA to match the magnitude of currents seen in recordings of CA1 PCs from young and aged animals, respectively \cite{campbell1996aging}.

@@

\subsection*{Modeling age-related changes in spike frequency adaptation}
Many CA1 PCs respond to square-pulse current injection by firing several early spikes followed by adaptation which slows the frequency of firing \cite{gant2006early,gu2008sk,moyer1992nimodipine}. To generate this firing pattern, we set the ionic currents to be the same amplitude range as observed in recordings of young adult CA1 PCs, with amplitudes for the {\Na} and {\Ca} currents at $\sim$2-3 nA, the DK current approximately double the inward cationic currents, and the SK current at $\sim$400-700 pA, depending on the intracellular {\Ca} concentration (see Tables \ref{tab:params} and \ref{tab:regimes}). This balance of ionic currents successfully generates adaptive firing similar to recordings (Fig.~\ref{fig:adapt}\edit{A} and \edit{\nameref{S2_Fig}}).

\begin{figure}[h!]
\centering
%\includegraphics[width=0.8\textwidth]{figures/Fig1.TIFF}
\caption{{\bf Adaptive firing in model PCs.}
\edit{(A.)} Adaptive firing in the yPC (solid black traces) versus aPC (dashed red traces) in response to a 800 ms \edit{150 pA} square-pulse stimulation \edit{shown in (E.)}. Corresponding {\Ca} currents, intracellular {\Ca} concentration, and SK currents are \edit{shown} in \edit{(B.), (C.), and (D.)}, respectively. \edit{Parameters for yPC: $a_{NaT}=1000$, $a_{CaL}=25$, $a_{DK}=8000$, $a_{SK}=1400$, $r_{w}=1.0$, $r_{c}=1e^{-3}$, and $k_{c}=3e^{-6}$. All parameters for aPC the same except $a_{CaL}=50$. Parameter units, and additional parameters kept constant for all simulations, are in Table \ref{tab:params}.}}
\label{fig:adapt}
\end{figure}

Studies show that adaptation is more pronounced in aged than young animal cells, leading to a shorter initial period of fast spiking, followed by fewer spikes or complete cessation of firing \cite{disterhoft1996calcium,gant2006early,moyer1992nimodipine,tombaugh2005slow}. To compare the young (yPC) and aged (aPC) model cells, all parameters were fixed except for the maximum amplitude of the L-type {\Ca} current, which was set to produce currents of $\sim$2-3~nA (young) or $\sim$5-6~nA (aged), based on recordings \cite{campbell1996aging}. This difference in the {\Ca} current causes the rate of firing in the first $\sim$100 ms (early firing) to decrease from 60 Hz in the yPC to 40 Hz in the aPC (Fig.~\ref{fig:adapt}\edit{A and \nameref{S4_Fig}A}). Firing for the remaining stimulation time (late firing) is also affected, decreasing from $\sim$6 Hz to $\sim$3 Hz in the yPC versus aPC, respectively \edit{(Fig.~\ref{fig:adapt}A and \nameref{S4_Fig}B}). The effect and frequencies are similar to those seen in recordings of CA1 PCs in young and old rabbits \cite{moyer1992nimodipine}.

Examining the {\Ca} and SK dynamics during the response reveals the mechanisms underlying the stronger adaptation in the aPC (Fig. \ref{fig:adapt}\edit{B-D}). The first two spikes occur nearly simultaneously in the two model cells. However, the larger increase in intracellular {\Ca} in the aPC induces a larger SK current, which in turn slows the cell's firing. The aPC falls behind the yPC by the third spike, and then slows its firing further as the response continues.

@@

AHP generation has been studied in CA1 PCs \cite{storm1989after}, particularly in the context of aging  \cite{disterhoft1996calcium,gant2006early,matthews2009fast,moyer1992nimodipine}. To induce AHPs, we kept the same parameters as in the previous simulations. We then stimulated model PCs with a 100 ms square pulse of sufficient amplitude to generate a burst of 4 APs (Fig. \ref{fig:AHP}). The AHPs produced under these conditions in the yPC have a peak amplitude of 3-4 mV \edit{(\nameref{S4_Fig} inset)}, similar to recordings \cite{matthews2009fast,power2002age}. The aPC required \edit{35} pA more current than the yPC to fire the same number of spikes \edit{(Fig. \ref{fig:AHP}E)}. However, the aPC fires earlier than the yPC due to its increased {\Ca} current (Fig. \ref{fig:AHP}\edit{A} inset). The aPC generates an AHP 1-2 mV larger than seen in the yPC (Fig. \ref{fig:AHP}\edit{B}), similar to the difference observed in recordings between young and aged cells \cite{moyer1992nimodipine,gant2006early,power2002age}. In the model, this larger AHP is due to an increased accumulation of {\Ca} in the aPC, which in turn produces a larger SK current (Fig. \ref{fig:AHP}\edit{C, D, respectively)}.

\begin{figure}[h!]
\centering
%\includegraphics[width=0.8\textwidth]{figures/Fig2.TIFF}
\caption{{\bf AHPs in model PCs.}
\edit{(A.)} Responses of the yPC (solid black traces) and aPC (dashed red traces) to 100 ms pulse. Voltage zoom in \edit{(B.)} shows AHPs in detail. Corresponding {\Ca} concentrations and SK currents are \edit{shown in (C.), and (D.)}, respectively. \edit{Parameters same as in Fig. \ref{fig:adapt}, except for the current stimulation duration and amplitude. Current amplitude adjusted to the minimum required (within 1 pA) to produce 4
spikes in each cell, i.e. 106 pA in yPC and 141 pA in aPC, as shown in (E.).}}
\label{fig:AHP}
\end{figure}

@@

Some CA1 PCs fire bursts instead of trains of spikes \cite{su2001extracellular}, especially in certain developmental periods \cite{chen2005transitional}. Burst firing can be generated in the model with several different parameter combinations. For the following simulations, we modified several of the current amplitudes and select kinetics, all within physiological limits (see Table \ref{tab:regimes}). Under this parameter regime, model PCs are silent at rest but burst if stimulated (Fig. \ref{fig:burstStims} \edit{and \nameref{S5_Fig}}).

\begin{figure}[h!]
\centering
%\includegraphics[width=0.8\textwidth]{figures/Fig3.TIFF}
\caption{{\bf Stimulated bursting in model PCs.}
Bursting in the yPC (solid black traces) and aPC (dashed red traces) models in response to 800 ms current injections of \edit{50-170 pA, as indicated in panels (A.)-(E}.). Parameters for the yPC: $a_{NaT}= 1300$, $a_{CaL}=25$, $a_{DK}=6000$, $a_{SK}=1600$, $r_{w}=1.8$, $r_{c}=5e^{-3}$, $k_{c}=6e^{-6}$. All parameters for the aPC the same except $a_{CaL}=50$.}
\label{fig:burstStims}
\end{figure}

To explore the effects of aging on bursting, we fixed all parameters except for the maximum {\Ca} current amplitude, as previously. We then stimulated the two model PCs with a series of square-pulse current injections of increasing amplitudes (50-170 pA) to compare their responses. Again, the larger {\Ca} current in the aPC causes it to fire either sooner or nearly simultaneously with the yPC shortly after stimulus onset in all simulations. However, the relative timing of the PCs' firing after the first burst depends on the stimulus amplitude (Fig. \ref{fig:burstStims}).

At lower stimulation amplitudes (50 and 80 pA; \edit{Fig. \ref{fig:burstStims}A and B)}, the aPC continues to burst sooner or nearly simultaneously, but always fires fewer spikes per burst than the yPC (2 versus 3, respectively). As the stimulation amplitude increases (110 pA; \edit{Fig. \ref{fig:burstStims}C}), the two PCs again fire nearly simultaneously at the onset. However, because the aPC fires fewer spikes per burst, it is able to recover sooner and burst before the yPC for four cycles. It is only towards the end of the stimulus that the larger AHP in the aPC eventually brings it into sync again with the yPC. Finally, at the highest stimulation amplitudes (140 and 170 pA; \edit{Fig. \ref{fig:burstStims}D and E}), both PCs show stronger adaptation, but the yPC maintains something akin to bursting, while the aPC `devolves' to a pattern more like tonic spiking. Depending on what is considered the relevant electrical event -- the single spike or the burst -- the aPC shows an increased number of events relative to the yPC (13 spikes versus 8 burst-like events, respectively), but the normal bursting pattern is lost.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Aging is a physiological process that is still poorly understood, especially with respect to effects on the brain. There are open questions about aging that are difficult to answer with an experimental approach. Underlying challenges include  the difficulty of recording \textit{in vivo} single cell and network activity simultaneously with submillisecond  resolution, and brain compensatory mechanisms triggered by genetic, pharmacologic, or behavioral manipulations. Mathematical modeling can help address some of these questions by allowing us to fix parameters that cannot be controlled experimentally and investigate neural activity under different conditions. We present a biophysical minimal model of CA1 pyramidal cells (PCs) based on general expressions for transmembrane ion transport derived from thermodynamical principles. The model allows directly varying the contribution of ion channels by changing their number. By analyzing the dynamics of the model, we find parameter ranges that reproduce the variability in electrical activity seen in PCs. In addition, increasing the L-type {\Ca} channel expression in the model reproduces age-related changes in electrical activity that are qualitatively and quantitatively similar to those observed in PCs from aged animals. We also make predictions about age-related changes in PC bursting activity that, to our knowledge, have not been reported previously. We conclude that the model's biophysical nature, flexibility, and computational simplicity make it a potentially powerful complement to experimental studies of aging.

@@

It is not well understood how changes in ion channel gene expression and hippocampal PC excitability may affect neuron responsiveness and microcircuit output. In part, this is due
to challenges inherent in performing the needed experiments. Single PCs are difficult
to access in intact animals where hippocampal microcircuit function is
preserved. It is also difficult to tease apart the influence of the many different neurophysiological factors that change during aging. Mathematical modeling provides a means to understand more about the effects of aging on hippocampal cellular excitability by controlling factors we cannot control experimentally.

@@

\subsection*{Model}
To simulate the electrical activity of CA1 PCs, we used an extended version of a two-dimensional, biophysical model previously developed and characterized by two of the present authors  \cite{herrera2012membranes,herrera2013relating,herrera2018thermodynamic}. The present model differs from our previous formulations in that it includes \Ca dynamics, i.e. it is three-dimensional, allowing for additional behaviors like bursting \cite{avron1993basic}. In addition, it is specially tuned by incorporating specific channel variants and corresponding experimental data from hippocampal PCs, as described below. The equations for the ionic currents are derived from first principles of thermodynamics. Previous modeling studies have shown that to reproduce firing behaviors such as spike frequency adaptation and bursting, the minimum number of variables is three \cite{hindmarsh1984model,avron1993basic}. 
In particular, {\Ca} dynamics are important for producing adaptation and burst firing in CA1 PCs (for review see \cite{mckiernan2017ca1}). 
The model dynamics are therefore described by three ordinary differential equations for the time-dependent changes in the transmembrane potential ($v$, in mV), the proportion of open {\K} channels ($w$ in [0,1]), and the intracellular {\Ca} concentration ($c$, in $\mu$M), respectively   \cite{herrera2018thermodynamic}. Based on a well-known relationship between voltage-dependent activation of delayed rectifier {\K} channels and inactivation of {\Na} channels, $w$ also represents the proportion of inactivated {\Na} channels \cite{rinzel1985excitation,avron1991minimal}.

It is assumed that the membrane potential changes due to currents produced by ions transported across the membrane. We take into account currents mediated by a voltage-gated inactivating {\Na} channel ($I_{NaT}$), voltage-gated L-type {\Ca} channel ($I_{CaL}$), voltage- and {\Ca}-gated {\K} channels ($I_{DK}$ and $I_{SK}$, respectively), and a {\Na}/{\K}-ATPase ($I_{NaK}$). We also incorporate a forcing term ($I_F$), which can be used to stimulate the model PC and is explained in more detail below. The time-dependent change in membrane potential can be written as
\begin{eqnarray}
\label{eq:dvdt}
C_m \partial_t v= I_{F} - I_{NaT}(v,w) - I_{CaL}(v,c) - I_{DK}(v,w) - I_{SK}(v,c) - I_{NaK}(v). 
\end{eqnarray}
with $\partial_t$ representing the instantaneous change with respect to time. $C_m$ (pF) is a constant representing the change in the density of charge around the membrane with respect to voltage, typically referred to as membrane capacitance in models based on electrical circuits  \cite{herrera2020nonequivalent}. Based on recordings from rat CA1 PCs, it is  assumed that $C_m=25$  pF \cite{groc2002vivo}.

@@

For the dynamics of intracellular {\Ca}, we assume recovery toward a steady state $c_{\infty}$ at a rate $r_c$, with increments caused by the {\Ca} current $I_{Ca}$ \cite{avron1991minimal},
\begin{eqnarray}
\partial_t{c} = r_{c}({c_{\infty} - c}) - \tilde{k}_{c} I_{CaL}(v,c).
\label{eq:dcdt}
\end{eqnarray}
The term $\tilde{k}_c$ in Eq.~\eqref{eq:dcdt} is a conversion factor ($\mu$M/pCoul) that accounts for the effect of {\Ca} flux across the membrane on the intracellular {\Ca} concentration.

The term $I_{F}$ represents a stimulus \textit{forcing} the membrane; that is, current from an electrode ($I_{Stim}$), or time-dependent fluctuations from the local field potential (LFP). LFP activity is simulated by replacing the term $I_{F}$ with a  time-dependent, Ornstein-Uhlenbeck (OU) process with amplitude $a_{F}(t)$ (pA). The mean is represented by  $\mu_{F}$ (pA) (drift term)
\cite{rudolph2003characterization} given by \cite{gillespie1996exact} 
\begin{eqnarray}
\label{eq:OU}
{a}_{F}(t+\delta) &=& a_{F}(t) \lrRound{1 -  \frac{\delta}{\tau_{F}}}  +
\lrSquare{ \mu_{F} \delta + \eta(t) ~\sqrt{d_{Stim} \delta}},
\end{eqnarray}
where $\delta$ is a small time step, $\tau_{F}$ is a relaxation time, and $\eta(t)$ is an independent
white noise process with zero-mean and unit standard deviation. In our simulations, the mean is set close to the rheobase for the model PCs ($\sim$50 pA). The process has a variance
$\sigma_{F}^2 = d_{F} \delta /2$ (pA), which means $d_{F}$ can be
approximated if an estimation of the variance of the current $a_{F}$ is
available \cite{rudolph2004method,destexhe2004novel}.

@@

The term $J_{F}$ (1/ms) is the input current $I_F$ (pA) divided by $v_T C_m$. After the change in variables and the normalization of the current amplitudes,  Eq.~\eqref{eq:dcdt} changes to 
\begin{eqnarray}
\label{eq:dcdt2}
\partial_t{c} = r_{c}({c_{\infty} - c}) - k_{c} A_{CaL}n_{\infty}(u
) \varphi_{Ca}(u,c),
\end{eqnarray}
where $k_c = \tilde{k}_c v_T C_m$.

@@

By exploring the model through parameter variations, we were able to find parameter sets that produced different firing patterns, such as adaptive firing, conditional bursting, and spontaneous bursting. The rationale behind finding parameters for qualitatively different firing patterns that emerge in a three-dimensional model similar to this one was established by Av-Ron and colleagues \cite{avron1993basic}. For example, they show that one way to observe transitions between adaptive firing and bursting is to vary the ratio of DK to \Ca-dependent \K channels. Therefore, in our study, we began by setting the model parameters within the base physiological range given by experimental recordings in PCs, and then took into consideration the different ratios of select parameters and their effects. Once we defined these different restricted parameter ranges, we then tuned semi-manually, i.e. using Python for loops to quickly run through a series of parameter values using knowledge of how increases or decreases in specific parameters should change the firing of cells (e.g., a larger SK current leads to greater inhibition of firing, lower firing frequencies, etc.). The specific PC firing patterns are described in more detail in the Results section, but the respective parameter sets are included in Table \ref{tab:regimes} for ease of comparison.

@@

\subsection*{Simulations}
All code was written in Python 3.7.4 and run on MacBook Pro laptops with 2.9 GHz Intel Core i5 processors. Simulations were performed using functions from the Python library NumPy \cite{numpyHarris2020}. Figures were produced with the Python library Matplotlib \cite{hunter2007matplotlib}. OU processes were simulated using the pyprocess package \cite{mondaca200PyProcess}.

\subsection*{Resource availability}
Resources generated by this study (code, figures, and manuscript files) are available via GitHub (\href{https://github.com/emckiernan/agingCA1}{github.com/emckiernan/agingCA1}) and archived via Zenodo (\href{https://doi.org/10.5281/zenodo.6788229}{doi.org/10.5281/zenodo.6788229}). To facilitate reuse, resources are shared under open licenses (see the license file in our GitHub repository). To promote reproducibility, Python code is embedded in a Jupyter notebook \cite{kluyver2016jupyter} that explains the code, how to use it, and how to generate all figures, including those in the Supporting Information.

\subsection*{Study design}
While aged cells display a number of biophysical changes, we focused on their {\Ca} channel expression. Aged CA1 PCs show an increase in the number of functional transmembrane L-type {\Ca} channels  \cite{herman1998up,thibault1996increase,veng2002regionally}. In particular, CA1 PCs from aged rats have increased expression of Ca$_v$1.2 at the plasma membrane \cite{nunez2014surface}. With these results in mind, we decided to simulate one aspect of aging by changing the number of Ca$_v$1.2 channels in our model membrane. We asked the question, is a change in Ca$_v$1.2 expression sufficient to reproduce the various changes in excitability, such as increased spike frequency adaptation, observed experimentally in aged CA1 PCs? In addition, CA1 PCs are known to burst \cite{mckiernan2017ca1}, but we are not aware of studies comparing their bursting patterns in young versus aged animals. Therefore, we used our model to also investigate the effects of altered Ca$_v$1.2 channel expression on bursting activity. In all the following simulations, once the models have been tuned to a specific firing pattern, young and aged model PCs (hereafter referred to as yPC and aPC, respectively) are identical with respect to every parameter except the maximum amplitude of their L-type {\Ca} current, which is set to produce currents of $\sim$2-3 nA or $\sim$5-6 nA to match the magnitude of currents seen in recordings of CA1 PCs from young and aged animals, respectively \cite{campbell1996aging}.

@@

\subsection*{Modeling age-related changes in spike frequency adaptation}
Many CA1 PCs respond to square-pulse current injection by firing several early spikes followed by adaptation which slows the frequency of firing \cite{gant2006early,gu2008sk,moyer1992nimodipine}. To generate this firing pattern, we set the ionic currents to be the same amplitude range as observed in recordings of young adult CA1 PCs, with amplitudes for the {\Na} and {\Ca} currents at $\sim$2-3 nA, the DK current approximately double the inward cationic currents, and the SK current at $\sim$400-700 pA, depending on the intracellular {\Ca} concentration (see Tables \ref{tab:params} and \ref{tab:regimes}). This balance of ionic currents successfully generates adaptive firing similar to recordings (Fig.~\ref{fig:adapt}A and \nameref{S2_Fig}).

\begin{figure}[h!]
\centering
%\includegraphics[width=0.8\textwidth]{figures/Fig1.TIFF}
\caption{{\bf Adaptive firing in model PCs.}
(A.) Adaptive firing in the yPC (solid black traces) versus aPC (dashed red traces) in response to a 800 ms 150 pA square-pulse stimulation shown in (E.). Corresponding {\Ca} currents, intracellular {\Ca} concentration, and SK currents are shown in (B.), (C.), and (D.), respectively. Parameters for yPC: $a_{NaT}=1000$, $a_{CaL}=25$, $a_{DK}=8000$, $a_{SK}=1400$, $r_{w}=1.0$, $r_{c}=1e^{-3}$, and $k_{c}=3e^{-6}$. All parameters for aPC the same except $a_{CaL}=50$. Parameter units, and additional parameters kept constant for all simulations, are in Table \ref{tab:params}.}
\label{fig:adapt}
\end{figure}

Studies show that adaptation is more pronounced in aged than young animal cells, leading to a shorter initial period of fast spiking, followed by fewer spikes or complete cessation of firing \cite{disterhoft1996calcium,gant2006early,moyer1992nimodipine,tombaugh2005slow}. To compare the young (yPC) and aged (aPC) model cells, all parameters were fixed except for the maximum amplitude of the L-type {\Ca} current, which was set to produce currents of $\sim$2-3~nA (young) or $\sim$5-6~nA (aged), based on recordings \cite{campbell1996aging}. This difference in the {\Ca} current causes the rate of firing in the first $\sim$100 ms (early firing) to decrease from 60 Hz in the yPC to 40 Hz in the aPC (Fig.~\ref{fig:adapt}A and \nameref{S4_Fig}A). Firing for the remaining stimulation time (late firing) is also affected, decreasing from $\sim$6 Hz to $\sim$3 Hz in the yPC versus aPC, respectively (Fig.~\ref{fig:adapt}A and \nameref{S4_Fig}B). The effect and frequencies are similar to those seen in recordings of CA1 PCs in young and old rabbits \cite{moyer1992nimodipine}.

Examining the {\Ca} and SK dynamics during the response reveals the mechanisms underlying the stronger adaptation in the aPC (Fig. \ref{fig:adapt}B-D). The first two spikes occur nearly simultaneously in the two model cells. However, the larger increase in intracellular {\Ca} in the aPC induces a larger SK current, which in turn slows the cell's firing. The aPC falls behind the yPC by the third spike, and then slows its firing further as the response continues.

@@

AHP generation has been studied in CA1 PCs \cite{storm1989after}, particularly in the context of aging  \cite{disterhoft1996calcium,gant2006early,matthews2009fast,moyer1992nimodipine}. To induce AHPs, we kept the same parameters as in the previous simulations. We then stimulated model PCs with a 100 ms square pulse of sufficient amplitude to generate a burst of 4 APs (Fig. \ref{fig:AHP}). The AHPs produced under these conditions in the yPC have a peak amplitude of 3-4 mV (\nameref{S5_Fig} inset), similar to recordings \cite{matthews2009fast,power2002age}. The aPC required 35 pA more current than the yPC to fire the same number of spikes (Fig. \ref{fig:AHP}E). However, the aPC fires earlier than the yPC due to its increased {\Ca} current (Fig. \ref{fig:AHP}A inset). The aPC generates an AHP 1-2 mV larger than seen in the yPC (Fig. \ref{fig:AHP}B), similar to the difference observed in recordings between young and aged cells \cite{moyer1992nimodipine,gant2006early,power2002age}. In the model, this larger AHP is due to an increased accumulation of {\Ca} in the aPC, which in turn produces a larger SK current (Fig. \ref{fig:AHP}C, D, respectively).

\begin{figure}[h!]
\centering
%\includegraphics[width=0.8\textwidth]{figures/Fig2.TIFF}
\caption{{\bf AHPs in model PCs.}
(A.) Responses of the yPC (solid black traces) and aPC (dashed red traces) to 100 ms pulse. Voltage zoom in (B.) shows AHPs in detail. Corresponding {\Ca} concentrations and SK currents are shown in (C.), and (D.), respectively. Parameters same as in Fig. \ref{fig:adapt}, except for the current stimulation duration and amplitude. Current amplitude adjusted to the minimum required (within 1 pA) to produce 4 spikes in each cell, i.e. 106 pA in yPC and 141 pA in aPC, as shown in (E.).}
\label{fig:AHP}
\end{figure}

@@

Some CA1 PCs fire bursts instead of trains of spikes \cite{su2001extracellular}, especially in certain developmental periods \cite{chen2005transitional}. Burst firing can be generated in the model with several different parameter combinations. For the following simulations, we modified several of the current amplitudes and select kinetics, all within physiological limits (see Table \ref{tab:regimes}). Under this parameter regime, model PCs are silent at rest but burst if stimulated (Fig. \ref{fig:burstStims} and \nameref{S6_Fig}).

\begin{figure}[h!]
\centering
%\includegraphics[width=0.8\textwidth]{figures/Fig3.TIFF}
\caption{{\bf Stimulated bursting in model PCs.}
Bursting in the yPC (solid black traces) and aPC (dashed red traces) models in response to 800 ms current injections of 50-170 pA, as indicated in panels (A.)-(E.). Parameters for the yPC: $a_{NaT}= 1300$, $a_{CaL}=25$, $a_{DK}=6000$, $a_{SK}=1600$, $r_{w}=1.8$, $r_{c}=5e^{-3}$, $k_{c}=6e^{-6}$. All parameters for the aPC the same except $a_{CaL}=50$.}
\label{fig:burstStims}
\end{figure}

To explore the effects of aging on bursting, we fixed all parameters except for the maximum {\Ca} current amplitude, as previously. We then stimulated the two model PCs with a series of square-pulse current injections of increasing amplitudes (50-170 pA) to compare their responses. Again, the larger {\Ca} current in the aPC causes it to fire either sooner or nearly simultaneously with the yPC shortly after stimulus onset in all simulations (\nameref{S8_Fig}). However, the relative timing of the PCs' firing after the first burst depends on the stimulus amplitude (Fig. \ref{fig:burstStims}).

At lower stimulation amplitudes (50 and 80 pA; Fig. \ref{fig:burstStims}A and B), the aPC continues to burst sooner or nearly simultaneously, but always fires fewer spikes per burst than the yPC (2 versus 3, respectively). As the stimulation amplitude increases (110 pA; Fig. \ref{fig:burstStims}C), the two PCs again fire nearly simultaneously at the onset. However, because the aPC fires fewer spikes per burst, it is able to recover sooner and burst before the yPC for four cycles. It is only towards the end of the stimulus that the larger AHP in the aPC eventually brings it into sync again with the yPC. Finally, at the highest stimulation amplitudes (140 and 170 pA; Fig. \ref{fig:burstStims}D and E), both PCs show stronger adaptation, but the yPC maintains something akin to bursting, while the aPC `devolves' to a pattern more like tonic spiking. Depending on what is considered the relevant electrical event -- the single spike or the burst -- the aPC shows an increased number of events relative to the yPC (13 spikes versus 8 burst-like events, respectively), but the normal bursting pattern is lost.
